---
title: "R Exercises"
author: "Sam Vincent"
date: "2025-09-19"
output:
  html_document: default
  pdf_document: default
---

# R Exercises

This will be used for recording various code blocks in R Studio.

## 1. Steps

This exercise contains code for calculating and manipulating pedometer data, using logical vectors.

```{r}
# Retrieving data and assigning it to variables
steps <- c(106, 0, 12775, 8287, 9222, 7080, 6055) 
kcal <- c(1356, 1341, 2109, 1882, 1970, 1938, 1851)

# Creating further calculations and assigning them to variables
daily_steps <- steps / 24
daily_energy <- kcal / daily_steps
hourly_steps <- daily_steps / 24

# Calculating total hourly steps and displaying the result
tot_hourly_steps <- sum(hourly_steps)
tot_hourly_steps
```

How to access specific values within this vector:

```{r}
steps[1] # Selecting one value
steps[1:5] # Selecting a range of values
steps[c(1,3,5)] # Selecting a non-continuous range of values
```

Introducing logic into our vectors:

```{r}
steps > 1500 # Are there more than 1500 steps?
steps < 1500 # Are there less than 1500 steps?

steps >= 1500 # Are there at least 1500 steps?
steps <= 1500 # Are there at most 1500 steps?

steps == 1500 # Are there exactly 1500 steps?
steps != 1500 # Are there any instances where there aren't 1500 steps?
```

Introducing advanced logic into our vectors:

```{r}
steps > 1000 & kcal < 2100 # When were more than 1000 steps taken AND less than 2100 kcals were burned?
steps <= 1200 | kcal != 1980 # When were at most 1200 steps taken OR anything other than 1980 kcals were burned?
```

Using logic to subset data:

```{r}
grand_steps <- steps > 1000 # Identifies days where steps > 1000
grand_kcal <- kcal[grand_steps] # Takes these days and extracts the corresponding kcal values

mean(grand_kcal) # Calculates the mean of all kcal values where steps > 1000
```

### Practice

Attempting to recalculate the mean energy consumption per step using only that part of the data that exceeded the BMR.

```{r}
# Reintroducing data
steps <- c(106, 0, 12775, 8287, 9222, 7080, 6055) 
kcal <- c(1356, 1341, 2109, 1882, 1970, 1938, 1851)

# Determining BMR
step0 <- steps == 0
BMR <- kcal[step0]
BMR

# Subtracting BMR value from kcal values
above_BMR <- (kcal - BMR)
above_BMR

# Calculating energy consumption per step
kcal_perstep <- (steps / kcal)
kcal_perstep
```

# T-tests & Wilcox on Existing Data

Using actual data for performing T and Wilcox tests

```{r}
# Initial data retrieval
library(dplyr)
df_SD <- read.csv("Q:/sampleData.csv")
df_SMD <- read.csv("Q:/sampleMetadata.csv")

# Data modification and display
df_SD
df_SMD
```

## Performing the T-tests

```{r}
# Assigning correponding values to vector variables
sampleid_untreated = df_SMD$SampleID[df_SMD$Treatment == "Untreated"]
sampleid_carbo = df_SMD$SampleID[df_SMD$Treatment == "Carboplatin 50mg/kg"]

# Removing the first column from df_SD
updated_data <- df_SD[,-1]
updated_data

# Performing T-Tests on each row of df_SD
ttest_results <- vector()
for(line in 1:nrow(updated_data)){
  ttest_results[line] = t.test(updated_data[line,colnames(updated_data) %in% sampleid_untreated],
          updated_data[line, colnames(updated_data) %in% sampleid_carbo])$p.value
  
}
ttest_results
```

## Performing the Wilcox Tests
```{r}
# Performing a wilcox test on all rows of df_SD
wilcox_results <- vector()
for(line in 1:nrow(updated_data)){
  wilcox_results[line] = wilcox.test(as.numeric(updated_data[line,colnames(updated_data) %in% sampleid_untreated]),
          as.numeric(updated_data[line, colnames(updated_data) %in% sampleid_carbo]))$p.value
  
}
wilcox_results
```

## Creating histograms based off the data
```{r}
hist(ttest_results, main = "T-Test P-Value Distribution Between\nUntreated and Carboplatin (50mg/kg) Mice", xlab = "P-Values", col = "lightblue", border = "black")
hist(wilcox_results, main = "Wilcox P-Vlaue Distribution Between\nUntreated and Carboplatin (50mg/kg) Mice", xlab = "P-Values", col = "lightgreen", border = "black")
```


